Medical Concept Extraction

### 2012-12-17 paper/paper.tex
typo fixes

\section{Background}
This section discusses a limited selection of previous work in concept extraction as well as work done in machine learning that is leveraged in this project.

\subsection{Related Work}
Concept extraction in medicine is an increasingly well-studied problem. One of the main driving forces behind such research is the annual i2b2/VA Workshop on Natural Language Processing Challenges for Clinical Records, which challenges participants to develop a system for "the extraction of medical concepts from patient reports." Torii et al. \cite{torii}, for example, developed a system using the same training data as us for the workshop. Their concept extraction system was trained using a tagger specifically developed for the biology domain, and their work shows the difficulty in porting trained systems across clinical documents from various sources.

In a work independent of the i2b2 workshop, Meystre and Haug \cite{meystre} developed a system for automatically retrieve medical problems from free-form text. Their system addresses one of the labels extracted from our system and references a pre-defined Problem List rather than training on data that is presumed to be a subset of possible terms. Tackling an interesting related problem, Happe et al. \cite{heppe} present a system for automatic concept extraction from $spoken$ medical reports in French, with robust performance in testing in spite of sensitivities to ambiguous pronunciations and acronyms, which are common in automatic speech recognition systems.

@@

Our system addresses the particular challenge of assigning concept labels to terms in medical records, but the task of assigning labels from multiple classes over a set of entries is a classic machine learning problem. Supervised multi-class classification algorithms assign labels to input data entries using a model trained on a data set of pre-labeled observations. There are countless methods that support multi-class classification--for the purpose of this work, we employ three well-known strategies (general Support Vector Machines, linear Support Vector Machines, and Conditional Random Fields) and compare their respective accuracies given our data sets and chosen features (section 4).

@@

Often problem sets are not linearly separable in feature space. To overcome this challenge, SVMs often map the original feature space into a higher-dimensional space--the specific mapping defined by a kernel function, which is chosen for efficiency based on the specific classification problem. Separating hyperplanes, which are defined as vectors, can then be chosen as linear combinations of feature vectors in the training set.

@@

\subsubsection{Conditional Random Fields}
Conditional random fields (CRF), unlike SVMs, is a probabilistic modeling method. The strength of CRF's in the case of classification for NLP is that it can take neighboring samples (i.e. context) into account when handling individual observations. A CRF is a type of undirected graphical model, which models a conditional probability distribution between observations and output values, which are represented as nodes. Given a sequence of words, this graph takes the form of a chain analogous to a Hidden Markov model. The conditional dependency of each output label on a given input is computed using a feature function in which features are given specific numerical weights and combine to determine the label's likelihood.

@@

\subsection{Biased Data}
Many machine learning classifiers assume relatively balanced data with respect to the class labels available. However, in the task of concept extraction it is most often the case that there is a null concept class indicating that the token is part of no relevant concept which is overwhelmingly more abundant than the concepts. The bias of the class labels in the data presents a challenge because a classifier optimized with respect to accuracy will be inclined to simply assign the null concept class for all labels since that will result in relatively high accuracy. Such classifiers will not, however, perform well with respect to more meaningful measures such as precision, recall, or F1-measure.

@@

\subsubsection{MIT Regular Expressions}
In order to capture some general characteristics of each word we included a set of regular expressions each of which adds a dimension to our classification. Such features help allow not only to pick up on traits within different medical categories--such as tests which are often abbreviated or contain numbers--but also to easily rule out non-medical terms such as days, phone numbers, etc. The characteristics we chose are as follows:
\begin{itemize}
\item {\tt INITCAP}: first letter is capitalized and alphabetic,
\item {\tt ALLCAPS}: entire word is capitalized and alphabetic,
\item {\tt CAPSMIX}: the word is a mix of capital and lowercase alphabetic characters,
\item {\tt HASDIGIT}: there is a digit in the word,
\item {\tt SINGLEDIGIT}: the token is a single digit,
\item {\tt DOUBLEDIGIT}: the token is a two digits,
\item {\tt FOURDIGITS}: the token is a series of four digits,
\item {\tt NATURALNUM}: the token is a natural number,
\item {\tt REALNUM}: the token is a real number,
\item {\tt ALPHANUM}: the token contains only digits and alphabetic characters,
\item {\tt HASDASH}: the token has a dash,
\item {\tt PUNCTUATION}: the token is punctuation,
\item {\tt PHONE1}: the token is a phone number,
\item {\tt PHONE2}: the token is a phone number with area code,
\item {\tt FIVEDIGIT}: the token is a series of five digits,
\item {\tt NOVOWELS}: the token does not contain vowels (i.e. abbreviations),
\item {\tt HASDASHNUMALPHA}: the token consists of a dash, numbers, and alphabetic characters, 
\item {\tt DATESEPERATOR}: the token is a date separator (e.g. {\tt [-/]})
\end{itemize}

\subsubsection{Porter Stemmer}
We already discussed the benefits of using entire words as features, but similar words come in many different forms. Motivated by generalization, we chose to use word stems as features and approached this feature extraction using three different stemming algorithms--two at the word-level and one at the sentence level (See 5.3.2). The addition of word stem features to our system improved our testing CRF F1 scores from 0.58 to 0.82.

Namely, we used the Porter Stemming Algorithm, as developed by Martin Porter.  This particular stemming algorithm is divided into a number of linear steps. Word letters are first replaced by symbols for vowels, V and consonants, C. The measure, m, of the word is computed as the number of repetitions of the pattern VC, and this is used to determine whether suffixes ought to be removed. Plural word forms, such as  sses $\rightarrow$ ss and removal of trailing s's, are then removed; followed by the removal of past participles; and finally terminal y's are replaced with i's. Subsequent steps handle different orderings of suffices by translating double suffices into a single suffix and then removing suffices under certain predefined conditions.

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same time. Additionally classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85.

@@

\item {\tt Metric Units}
A clear trend we noticed our medical record data was that numbers followed by metric weights (i.e. mg, milligrams, grams) are almost always part of treatment descriptions. Whereas numbers followed by metric lengths or areas are typically part of problem observations. We still believe this could be a useful feature for related systems.

@@

\item {\tt Flag Words}
Certain terms such as those described in 5.2.1 are consistently labeled as "problems", "tests", or "treatments". We attempted to create small dictionaries of these terms and assign flags to words that fall into one of the three dictionaries. The issue with this strategy is that words that words that should be labeled as one of our three medical terms but do not appear in the dictionary are now more closely aligned with purely non-medical words. That said, this feature improved accuracy for our three categories, but also yielded far too many false positives.

@@

\subsubsection{WordNet Lemmatizer}
Unlike the Porter Stemmer and the Lancaster Stemmer, which consider words independent of context, the WordNet Lemmatizer requires words to be processed at the sentence level in order to extract their root form. The WordNet Lemmatizer uses the part of speech tags for individual words, which are assign via NLTK as previously mentioned, in conjunction with a subset of tags used for morphology to extract stems from each word. WordNet references a database of word stems, and uses two types of processes to try to convert input strings into a form present in the database. A list of inflectional endings based on the word's syntactic tag is referenced in search of a match that can be detached from the word. An exception list is also referenced in search for an inflected form. WordNet first checks for exceptions and then uses rules for detachment to transform the word--after each transformation the result is referenced against the stem database until a match is eventually found.

@@

\section{System}
This section discusses the framework code we have developed in order to perform concept extraction, the web service frontend which allows users to test the service without seeing any of the underlying data, and the machine learning libraries that are leveraged.

@@

\section{Results}
This section discusses the data set, the models that were trained as part of the project, the (excessive) amount of time taken to train them, and the performance of those models on the test data.

@@

\begin{figure*}
\begin{center}
	\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l||l|l|}
	\hline
	\multicolumn{11}{|c||}{\bf Feature Set} & \multicolumn{2}{|c|}{\bf F1 Score}	\\ \hline
	% BEGIN COLUMN HEADERS
		\begin{sideways} length \end{sideways} & 
		\begin{sideways} mitre \end{sideways} & 
		\begin{sideways} next \end{sideways} & 
		\begin{sideways} pos \end{sideways} & 
		\begin{sideways} prev \end{sideways} & 
		\begin{sideways} stem\_lancaster \end{sideways} & 
		\begin{sideways} stem\_porter \end{sideways} & 
		\begin{sideways} stem\_wordnet \end{sideways} & 
		\begin{sideways} test\_result \end{sideways} & 
		\begin{sideways} word \end{sideways} & 
		\begin{sideways} word\_shape \end{sideways} & 
		LIN & CRF	\\ \hline
	% END COLUMN HEAERS

@@

\end{tabular}
\end{center}
\caption{A summary of the F1-scores from many different models with different features enabled.}
\label{fig:summary_f1_scores}
\end{figure*}

>>>>

\section{Background}
\subsection{Related Work}
Concept extraction in medicine is an increasingly well-studied problem. One of the main driving forces behind such research is the annual i2b2/VA Workshop on Natural Language Processing Challenges for Clinical Records, which challenges participants to develop a system for "the extraction of medical concepts from patient reports." Torii et al. \cite{torii}, for example, developed a system using the same training data as us for the workshop. Their concept extraction system was trained using a tagger specifically developed for the biology domain, and their work shows the difficulty in porting trained systems across clinical documents from various sources.

In a work independent of the i2b2 workshop, Meystre and Haug \cite{meystre} developed a system for automatically retrieving medical problems from free-form text. Their system addresses one of the labels extracted from our system and references a pre-defined Problem List rather than training on data that is presumed to be a subset of all possible terms. Tackling an interesting related problem, Happe et al. \cite{heppe} present a system for automatic concept extraction from $spoken$ medical reports in French, with robust performance in testing in spite of sensitivities to ambiguous pronunciations and acronyms, which are common in automatic speech recognition systems.

@@

Our system addresses the particular challenge of assigning concept labels to terms in medical records, but the task of assigning labels from multiple classes over a set of entries is a classic machine learning problem. Supervised multi-class classification algorithms assign labels to input data entries using a model trained on a data set of pre-labeled observations. There are countless methods that support multi-class classification--for the purpose of this work, we employ three well-known strategies (general Support Vector Machines, linear Support Vector Machines, and Conditional Random Fields) and compare their respective accuracies given our data sets and chosen features (section 4.2).

@@

Often problem sets are not linearly separable in feature space. To overcome this challenge, SVMs often map the original feature space into a higher-dimensional space--the specific mapping is defined by a kernel function, which is chosen for efficiency based on the specific classification problem. Separating hyperplanes, which are defined as vectors, can then be chosen as linear combinations of feature vectors in the training set.

@@

\subsubsection{Conditional Random Fields}
Conditional random fields (CRFs), unlike SVMs, use a probabilistic modeling method. The strength of CRFs in the case of classification for NLP is that it can take neighboring samples (i.e. context) into account when handling individual observations. A CRF is a type of undirected graphical model, which models a conditional probability distribution between observations and output values, which are represented as nodes. Given a sequence of words, this graph takes the form of a chain analogous to a Hidden Markov model. The conditional dependency of each output label on a given input node is computed using a feature function in which features are given specific numerical weights and combine to determine the label's likelihood.

@@

\subsection{Biased Data}
Many machine learning classifiers assume relatively balanced data with respect to the class labels available. However, in the task of concept extraction it is most often the case that there is a null concept class indicating that the token is part of no relevant concept which is overwhelmingly more abundant than the concepts. The bias of the class labels in the data presents a challenge, because a classifier optimized with respect to accuracy will be inclined to simply assign the null concept class for all labels (since that will result in relatively high accuracy). Such classifiers will not, however, perform well with respect to more meaningful measures such as precision, recall, or F1-measure.

@@

\subsubsection{MIT Regular Expressions}
In order to capture some general characteristics of each word we included a set of regular expressions each of which adds a dimension to our classification. Such features help not only to pick up on traits within different medical categories--such as tests which are often abbreviated or contain numbers--but also to easily rule out non-medical terms such as days, phone numbers, etc. The characteristics we chose are as follows:
\begin{itemize}
\item {\tt INITCAP}: first letter is capitalized and alphabetic,
\item {\tt ALLCAPS}: entire word is capitalized and alphabetic,
\item {\tt CAPSMIX}: the word is a mix of capital and lowercase alphabetic characters,
\item {\tt HASDIGIT}: there is a digit in the word,
\item {\tt SINGLEDIGIT}: the token is a single digit,
\item {\tt DOUBLEDIGIT}: the token is a two digits,
\item {\tt FOURDIGITS}: the token is a series of four digits,
\item {\tt NATURALNUM}: the token is a natural number,
\item {\tt REALNUM}: the token is a real number,
\item {\tt ALPHANUM}: the token contains only digits and alphabetic characters,
\item {\tt HASDASH}: the token has a dash,
\item {\tt PUNCTUATION}: the token is punctuation,
\item {\tt PHONE1}: the token is a phone number,
\item {\tt PHONE2}: the token is a phone number with area code,
\item {\tt FIVEDIGIT}: the token is a series of five digits,
\item {\tt NOVOWELS}: the token does not contain vowels (i.e. abbreviations),
\item {\tt HASDASHNUMALPHA}: the token consists of a dash, numbers, and alphabetic characters, 
\item {\tt DATESEPERATOR}: the token is a date separator (e.g. {\tt [-/]})
\end{itemize}

\subsubsection{Porter Stemmer}
We already discussed the benefits of using entire words as features, but similar words come in many different forms. Motivated by generalization, we chose to use word stems as features and approached this feature extraction using three different stemming algorithms--two at the word-level and one at the sentence level (See 5.3.2). The addition of word stem features to our system improved our testing CRF F1 scores from 0.58 to 0.82. We used built-in NLTK implementations of all of the stemming algorithms described in future sections.

Namely, we used the Porter Stemming Algorithm, as developed by Martin Porter.  This particular stemming algorithm is divided into a number of linear steps. Word letters are first replaced by symbols for vowels, V and consonants, C. The measure, m, of the word is computed as the number of repetitions of the pattern VC, and this is used to determine whether suffixes ought to be removed. Plural word forms are then removed such as  sses $\rightarrow$ ss and removal of trailing s's; followed by the removal of past participles; and finally terminal y's are replaced with i's. Subsequent steps handle different orderings of suffices by translating double suffices into a single suffix and then removing suffices under certain predefined conditions.

@@

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same type. Additional classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85.

@@

\item {\tt Metric Units}
A clear trend we noticed in our medical record data was that numbers followed by metric weights (i.e. mg, milligrams, grams) are almost always part of treatment descriptions. Whereas numbers followed by metric lengths or areas are typically part of problem observations. We still believe this could be a useful feature for related systems.

@@

\item {\tt Flag Words}
Certain terms such as those described in 5.2.1 are consistently labeled as "problems", "tests", or "treatments". We attempted to create small dictionaries of these terms and assign flags to words that fall into one of the three dictionaries. The issue with this strategy is that words that should be labeled as one of our three medical terms but do not appear in the dictionary are now more closely aligned with purely non-medical words. That said, this feature improved accuracy for our three categories, but also yielded far too many false positives.

@@

\subsubsection{WordNet Lemmatizer}
Unlike the Porter Stemmer and the Lancaster Stemmer, which consider words independent of context, the WordNet Lemmatizer requires words to be processed at the sentence level in order to extract their root form. The WordNet Lemmatizer uses the part of speech tags for individual words, which are assign via NLTK as previously mentioned, in conjunction with a subset of tags used for morphology to extract stems from each word. WordNet references a database of word stems, and uses two types of processes to try to convert input strings into a form present in the database: A list of inflectional endings based on the word's syntactic tag is referenced in search of a match that can be detached from the word. An exception list is also referenced in search for an inflected form. WordNet first checks for exceptions and then uses rules for detachment to transform the word--after each transformation the result is referenced against the stem database until a match is eventually found.

@@

\section{System}

@@

\section{Results}

@@

\begin{figure}
\begin{center}
	\begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l||l|l|}
	\hline
	\multicolumn{11}{|c||}{\bf Feature Set} & \multicolumn{2}{|c|}{\bf F1 Score}	\\ \hline
	% BEGIN COLUMN HEADERS
		\begin{sideways} length \end{sideways} & 
		\begin{sideways} mitre \end{sideways} & 
		\begin{sideways} next \end{sideways} & 
		\begin{sideways} pos \end{sideways} & 
		\begin{sideways} prev \end{sideways} & 
		\begin{sideways} stem\_lancaster \end{sideways} & 
		\begin{sideways} stem\_porter \end{sideways} & 
		\begin{sideways} stem\_wordnet \end{sideways} & 
		\begin{sideways} test\_result \end{sideways} & 
		\begin{sideways} word \end{sideways} & 
		\begin{sideways} word\_shape \end{sideways} & 
		LIN & CRF	\\ \hline
	% END COLUMN HEAERS

@@

\end{tabular}
\end{center}
\caption{A summary of the F1-scores from many different models with different features enabled.}
\label{fig:summary_f1_scores}
\end{figure}
